\section{Conclusion}

\subsection{Summary of Contributions}

This project designed and implemented \textbf{Geo-SLM}, a hybrid neuro-symbolic system
for extracting structured numerical data from chart images. The key contributions are:

\begin{enumerate}
    \item \textbf{Hybrid 5-stage pipeline architecture}: A modular pipeline combining
          deep learning (YOLO, EfficientNet-B0), vision-language model extraction
          (DePlot/MatCha/Pix2Struct/SVLM), and AI reasoning (SLM with fallback to
          cloud APIs). Each stage operates on validated Pydantic schema contracts,
          enabling independent development, testing, and replacement
          (demonstrated by the Stage~3 VLM rewrite replacing 1,109 lines of geometry
          code with a 309-line pluggable extractor without modifying any downstream stage).

    \item \textbf{VLM-based chart data extraction}: A pluggable 4-backend VLM extraction
          architecture (DePlot, MatCha, Pix2Struct-base, SVLM/Qwen2-VL) that converts
          chart images directly into structured tables via image-to-text derendering.
          A 50-chart benchmark with Gemini Vision ground truth confirmed that the prior
          geometry approach failed completely (0\% axis detection, 0 OCR output), directly
          motivating this design. The VLM approach achieves 100\% extraction completion
          across all 50 benchmark charts.

    \item \textbf{Large-scale academic chart dataset}: 32,364 classified charts from
          $\sim$4,000 arXiv papers. The derived SLM training dataset contains 268,799
          ChatML conversation samples across 8 chart types, generated using the
          geometry-era Stage~3 pipeline (v1--v5). VLM-quality re-extraction is planned
          for the next training iteration.

    \item \textbf{Language-extensible architecture for Vietnamese chart Q\&A}: The
          pipeline is designed with a Core-first, Localize-second strategy where
          6 out of 11 components are fully language-agnostic. Vietnamese support is
          achieved through the SVLM (Qwen2-VL) backend which supports multilingual
          input, combined with targeted data generation (Vietnamese QA samples)
          -- without modifying any core pipeline logic.

    \item \textbf{Comprehensive evaluation}: Chart detection at 93.5\% mAP@0.5,
          classification at 94.14\% accuracy, and 100\% batch extraction success rate
          across the full 32,364-chart corpus with 92.6\% overall pipeline confidence.
\end{enumerate}

\subsection{Achievement of Objectives}

\begin{table}[htbp]
\centering
\caption{Achievement of project objectives.}
\label{tab:objectives}
\begin{tabular}{@{}lllc@{}}
\toprule
\textbf{Objective} & \textbf{Target} & \textbf{Achieved} & \textbf{Status} \\
\midrule
Detection mAP@0.5 & $>85\%$ & 93.5\% & Exceeded \\
Classification accuracy & $>90\%$ & 94.14\% & Exceeded \\
Extraction completion (VLM) & 100\% & 100\% & Met \\
Training dataset size & $>200{,}000$ & 268,799 & Exceeded \\
Consumer GPU deployment & RTX 3060 & Confirmed & Met \\
Vietnamese Q\&A support & End-to-end & Architecture ready & In progress \\
SLM fine-tuning & Complete & In progress & Pending \\
Public benchmark eval & ChartQA/PlotQA & Not yet & Pending \\
\bottomrule
\end{tabular}
\end{table}

Five of seven objectives were exceeded. The Vietnamese Q\&A architecture is in place
(12/17 components language-agnostic, localization layers designed), pending QA data
generation and SLM fine-tuning. The remaining items are in progress.

\subsection{Future Work}

\begin{enumerate}
    \item \textbf{Vietnamese chart Q\&A deployment}: Execute the full Vietnamese
          localization plan: configure OCR for Vietnamese diacritical marks, generate
          5,000--10,000 Vietnamese QA samples via Data Factory, fine-tune SLM with
          combined English + Vietnamese data, and evaluate Vietnamese chart understanding
          accuracy end-to-end.

    \item \textbf{SLM fine-tuning and evaluation}: Complete QLoRA training of
          Qwen-2.5-1.5B on the 268,799-sample dataset (English) plus Vietnamese
          supplement. Target: JSON valid rate $>$95\%, field accuracy $>$90\%,
          Vietnamese output fluency $>$90\%, latency $<$2s per chart.

    \item \textbf{Public benchmark comparison}: Evaluate on ChartQA
          \cite{masry2022chartqa} and PlotQA \cite{methani2020plotqa} for direct
          comparison with DePlot \cite{liu2023deplot} and MatCha \cite{lee2023matcha}.

    \item \textbf{VLM backend ablation study}: Systematically measure the contribution
          of each VLM backend (DePlot vs MatCha vs Pix2Struct-base vs SVLM) on the
          50-chart benchmark to formally select the optimal default backend and
          characterize trade-offs between accuracy, latency, and Vietnamese support.

    \item \textbf{VLM inference optimization}: Implement quantized VLM inference
          (INT8/INT4) and batch processing to reduce Stage~3 latency below 2~s
          per chart from the current 2--4~s baseline.

    \item \textbf{SLM training on VLM-extracted features}: Re-run Stage~3 batch
          extraction with VLM backends on the full 32,364-chart corpus to generate
          VLM-quality training features, replacing the geometry-era v3 dataset.

    \item \textbf{Vietnamese domain expansion}: Collect chart images from Vietnamese
          government reports, business dashboards, and academic publications to build
          a Vietnamese chart benchmark dataset.

    \item \textbf{Production deployment}: Integrate the Celery task queue and FastAPI
          server for asynchronous pipeline execution with PostgreSQL state management.

    \item \textbf{Dataset release}: Prepare the training dataset (English + Vietnamese)
          and model weights for public release to support community research.
\end{enumerate}